---
title: 'SDM Benchmark Study Part 7: Fitting and Testing Inhomogeneous Poisson Process Models with Updated Test Data'
author: "Benton Tripp"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    df_print: paged
  
---

<style>
.tocify-extend-page {
  height: 0 !important;
}
.post_navi {
  display: flex
}

.post_navi-label {
  font-size: 0.8em;
  opacity: 0.5
}

.post_navi .post_navi-item {
  padding: 0 2.2em;
  width: 50%;
  position: relative;
  color: inherit !important
}

.post_navi .nav_prev {
  text-align: left
}

.post_navi .nav_next {
  text-align: right
}

.post_navi .nav_prev .post_navi-arrow {
  left: 0
}

.post_navi .nav_next .post_navi-arrow {
  right: 0
}

.post_navi .post_navi-arrow {
  position: absolute;
  top: 50%;
  transform: translateY(-50%);
  font-size: 2.5em;
  opacity: 0.3
}

footer .wrapper {
  max-width: -webkit-calc(800px - (30px*2));
  max-width: calc(800px - (30px*2));
  margin-right: auto;
  margin-left: auto;
  padding-right: 30px;
  padding-left: 30px
}

@media screen and (max-width:800px) {
  footer .wrapper {
    max-width: -webkit-calc(800px - (30px));
    max-width: calc(800px - (30px));
    padding-right: 15px;
    padding-left: 15px
  }
}

footer .wrapper:after,
.footer-col-wrapper:after {
  content: "";
  display: table;
  clear: both
}

.svg-icon {
  width: 26px;
  height: 16px;
  display: inline-block;
  fill: #828282;
  padding-right: 5px;
  vertical-align: text-top
}

.social-media-list li+li {
  padding-top: 5px
}

.site-footer {
  border-top: 1px solid #e8e8e8;
  padding: 30px 0
}

.footer-heading {
  font-size: 18px;
  margin-bottom: 15px
}

.contact-list,
.social-media-list {
  list-style: none;
  margin-left: 0;
  width: 155px;
}

.footer-col-wrapper {
  font-size: 15px;
  color: #828282;
  margin-left: -15px
}

.footer-col {
  float: left;
  margin-bottom: 15px;
  padding-left: 15px
}

.footer-col-1 {
  width: 30%;
}

.footer-col-2 {
  width: 25%;
}

.footer-col-3 {
  width: 45%;
}

@media screen and (max-width:800px) {

  .footer-col-1,
  .footer-col-2 {
    width: -webkit-calc(50% - (30px/2));
    width: calc(50% - (30px/2))
  }

  .footer-col-3 {
    width: -webkit-calc(100% - (30px/2));
    width: calc(100% - (30px/2))
  }
}

@media screen and (max-width:600px) {
  .footer-col {
    float: none;
    width: -webkit-calc(100% - (30px/2));
    width: calc(100% - (30px/2))
  }
}
</style>

```{r setup, include=F, warning=F, message=F}
knitr::opts_chunk$set(echo = T, message=F, warning=F, cache=F, root.dir="..")
```

## Setup

```{r setup-2, results='hide'}

library(sf)
library(terra)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(readr)
library(data.table)
library(knitr)
library(purrr)
library(caret)
library(spatstat)
library(tidyr)

# Set seed for splitting and modeling
set.seed(19)

# Load the dataset saved in part 2 of the study 
df <- readRDS("artifacts/final_data/final_data.rds") %>% setDT()

# Define some global variables that will be referenced throughout the modeling 
states <- sort(unique(df$state))
species <- sort(unique(df$common.name))
spec.state <- expand.grid(species=species, 
                          state=states, 
                          stringsAsFactors=F)

# Get model or other object from cache if it has been saved before
get.object <- function(obj, file.name, obj.path, read.func=readRDS, save.func=saveRDS, ...) {
  f.path <- file.path(obj.path, file.name)
  if (!dir.exists(obj.path)) {
    dir.create(obj.path)
  }
  # Object cache check
  if (file.exists(f.path)) {
    obj <- read.func(f.path)
  } else {
    save.func(obj, f.path, ...)
  }
  obj
}

```

## IPP Models

### Load Variable Importance

```{r varimp}
# Load variable importance from fitted LASSO models
lasso.model.path="artifacts/models/lasso_3_fs"

# Define min/max scaling function for rasters
min.max.scale <- function(x, na.rm=T) {
  min.x <- min(x, na.rm=na.rm)
  max.x <- max(x, na.rm=na.rm)
  (x - min.x) / (max.x - min.x)
}

get.var.imp <- function(st, spec, dir="artifacts/models/lasso_3_fs") {
  files <- list.files(dir) %>%
    .[grepl(paste(tolower(gsub(" ", "_", spec)), st, sep="_"), .)] %>%
    file.path(dir, .)
  if (length(files) == 0) return(
    tibble(
      common.name=character(0),
      state=character(0),
      variable=character(0),
      importance=numeric(0),
      wt=numeric(0),
      weighted.imp=numeric(0)
    )
  )
  var.imp <- purrr::map_df(files, ~{
    fit <- readRDS(.x)
    coef.df <- coef(fit$finalModel, s = fit$bestTune$lambda) %>%
      as.matrix() %>%
      as.data.frame()
    # Remove the intercept
    coef.df <- coef.df[-1, , drop = F]
    # Create a data frame of variable importance
    var.importance <- tibble(
      common.name = spec,
      state = st,
      variable = rownames(coef.df),
      importance = abs(coef.df[,1])
    ) %>%
      # Rank variables by importance
      arrange(state, common.name, -importance, variable) %>%
      # Only look at variables where imp. > 0
      filter(importance > 0)
  }) %>%
    mutate(n=n()) %>%
    group_by(common.name, state, variable, n, .drop=F) %>%
    summarize(importance=median(importance), 
              wt=n()) %>%
    ungroup() %>%
    mutate(wt=wt/n) %>%
    select(-n) %>%
    mutate(weighted.imp=min.max.scale(wt*importance)) %>%
    arrange(-weighted.imp, variable) 
}

var.imp <- purrr::map_df(1:nrow(spec.state), function(i) {
  spec <- spec.state[i,]$species
  st <- spec.state[i,]$state
  get.var.imp(st, spec)
})

var.imp

```


```{r ipp-model, results='hide'}

get.f1 <- function(locations.df, thresh=0.5, method="F1") {
  metrics <- locations.df[, .(pred=factor(ifelse(p.obs > thresh, T, F), 
                               levels=c("FALSE", "TRUE")), 
          obs=factor(obs, levels=c("FALSE", "TRUE")))] %>% 
    confusionMatrix(reference=.$pred, data=.$obs, 
                    positive = "TRUE", 
                    mode="everything") %>%
    as.list() %>%
    .[["byClass"]] %>%
    as.list()
  if (method == "F1") {
    return(metrics$F1)
  } else if (method == "SS") {
    # Penalize heavily if either is near zero or NA
    if (is.na(metrics$Sensitivity) || is.na(metrics$Specificity)) {
      return(-1e6) 
    } else if (metrics$Sensitivity < .05 || metrics$Specificity < .05) {
      return(-1e6) 
    }
    
    # Combine Specificity and Sensitivity (e.g., geometric mean)
    combined.metric = sqrt(metrics$Sensitivity * metrics$Specificity)
    return(combined.metric)
  }
}

optimize.f1 <- function(locations.df) {
  objective.fn <- function(thresh) {
    # Negative because we want to maximize
    -get.f1(locations.df=locations.df, thresh=thresh, method="SS")  
  }
  opt.result <- optimize(objective.fn, lower=0.05, upper=0.95)
  return(opt.result$minimum)  # Return the optimal threshold
}

get.acc <- function(test, thresh) {
  df <- test[, .(pred=as.factor(ifelse(p.obs > thresh, T, F)), obs=as.factor(obs))]
  cm <- confusionMatrix(df$pred, df$obs, positive = "TRUE", mode="everything")
}

# 9, 26, 27, 32
purrr::walk(1:nrow(spec.state), function(i) {
  covariates.keep <- 50
  spec <- spec.state[i,]$species
  st <- spec.state[i,]$state
  glm.results.path <- file.path("artifacts/test_results/ipp_glm_mpl_2",
                                paste0(spec, "_", st, "_ipp_glm_mpl_2.rds"))
  if (!file.exists(glm.results.path)) {
    cat("Fitting IPP model for", spec, "in", st, "\n")
    
    cat("\tGetting pre-processed `spatstat.geom::ppp` object (train)...\n")
    Q <- readRDS(file.path("artifacts", "train_spatstat_Q_2",
                           paste0(st, "_", spec, "_Q.rds")))
    cat("\tGetting `spatstat.geom::ppp` object (test)...\n")
    Q.test <- readRDS(file.path("artifacts", "test_spatstat_Q_2",
                                paste0(st, "_", spec, "_Q.rds")))
    
    converged <- F
    no.pred.error <- F
    # roc.obj <- NULL
    spec.sens.check <- F
    alternate.formula <- F
    while (!converged | !no.pred.error | !spec.sens.check) {
      # Set path
      glm.path <- file.path("artifacts/models/ipp_glm_mpl_2", 
                            paste0(spec, "_", st, "_", covariates.keep,
                                   "_ipp_glm_mpl_2.rds"))
      if (covariates.keep > 0) {
        # Select covariates based on feature importance
        cat("\tFetching variable importance with `covariates.keep` set to", 
            covariates.keep, "\n")
        fs.df <- var.imp %>% 
          filter(state == st & common.name == spec) %>%
          mutate(var1 = purrr::map_chr(variable, 
                                       ~ stringr::str_split(.x, "\\:")[[1]][1]),
                 var2 = purrr::map_chr(variable, ~ {
                   split_result <- stringr::str_split(.x, "\\:")[[1]]
                   if(length(split_result) > 1) split_result[2] else NA_character_
                 })) %>%
          mutate(variable = ifelse(is.na(var2), 
                                   var1, 
                                   paste(var1, var2, sep = ":"))) %>%
          # Keep only pre-determined # of variables/interactions
          head(covariates.keep)
        
        if (nrow(fs.df) > 0) {
          covariates <- c(fs.df$var1, fs.df$var2) %>% 
            unique() %>% 
            sort()
          
          # Load/compute filtered & pre-processed rasters
          covariates <- covariates %>%
            set_names() %>%
            purrr::map(function(.x) {
              file <- file.path("artifacts/spatstat_imgs", 
                                paste0(st, "_", .x, ".rds"))
              readRDS(file)
            })
          
          # Create formula
          if (alternate.formula) {
            if (length(covariates) < covariates.keep) {
              covariates.keep <- length(covariates) # in case there are fewer available
            } else if (length(covariates) > covariates.keep) {
              covariates <- covariates[1:covariates.keep]
            }
            
            # No interactions
            .f <- paste(names(covariates), collapse = " + ") %>% 
              paste("~", .) %>% 
              as.formula()
          } else {
            if (nrow(fs.df) < covariates.keep) {
              covariates.keep <- nrow(fs.df) # in case there are fewer available
            }
            # With interactions
            .f <- paste(fs.df$variable, collapse=" + ") %>% 
              paste("~", .) %>% 
              as.formula()
          }
          # Make sure glm.path is right
          glm.path <- file.path("artifacts/models/ipp_glm_mpl_2", 
                            paste0(spec, "_", st, "_", covariates.keep,
                                   "_ipp_glm_mpl_2.rds"))
          
        } else {
          cat("\tThere are no specified covariates for", spec, st, "\n")
        }
        
        # Fit the IPP model, using the Method of Maximum PseudoLikelihood (MPL)
        
        # GLM Model
        cat("\tFitting GLM Model using the Method of Maximum",
            "PseudoLikelihood...\n")
        fit.glm <- tryCatch({
          ppm(Q=Q, trend=.f, covariates=covariates, 
              rbord=.05, method="mpl", emend=T) %>%
            get.object(
              obj=.,
              file.name=paste0(spec, "_", st, "_", covariates.keep,
                               "_ipp_glm_mpl_2.rds"), 
              obj.path="artifacts/models/ipp_glm_mpl_2")}, 
          error=function(e) NULL)
        if (is.null(fit.glm)) {
          if (covariates.keep > 15) {
            covariates.keep <- covariates.keep - 5
          } else {
            covariates.keep <- covariates.keep - 1
          }
          if (covariates.keep < 3 & !alternate.formula) {
            cat("\tNo trend identified. Removing interactions...\n")
            alternate.formula <- T
            covariates.keep <- 50
          }  else if (covariates.keep < 1 & alternate.formula) {
            cat("\tNo trend identified. Removing covariates...\n")
          } else if (covariates.keep < 0 & alternate.formula) {
            stop("\tUnable to successfully fit a model given the data.\n")
          } 
          next
        }
        converged <- fit.glm$internal$glmfit$converged
      } else {
        cat("\tFitting GLM Model using the Method of Maximum",
            "PseudoLikelihood with no trend...\n")
        fit.glm <- ppm(Q=Q, rbord=.05, method="mpl", emend=T) %>%
          get.object(
            obj=.,
            file.name=paste0(spec, "_", st, "_", covariates.keep,
                             "_ipp_glm_mpl_2.rds"), 
            obj.path="artifacts/models/ipp_glm_mpl_2")
        converged <- T
      } 
      # Check for errors when predicting
      options(show.error.messages = F)
      no.pred.error <- suppressWarnings(
        tryCatch(
          expr={predict.ppm(fit.glm, se=T); T}, 
          error=function(e) {F}
        )
      )
      options(show.error.messages = T)
      if (!converged | !no.pred.error) {
        cat("\tThe model converged or there was an error",
                "for covariates.keep ==", covariates.keep, "\n")
        
        file.remove(glm.path)
        if (covariates.keep > 15) {
          covariates.keep <- covariates.keep - 5
        } else {
          covariates.keep <- covariates.keep - 1
        }
        if (covariates.keep < 3 & !alternate.formula) {
          cat("\tNo trend identified. Removing interactions...\n")
          alternate.formula <- T
          covariates.keep <- 50
        }  else if (covariates.keep < 1 & alternate.formula) {
          cat("\tNo trend identified. Removing covariates...\n")
        } else if (covariates.keep < 0 & alternate.formula) {
          stop("\tUnable to successfully fit a model given the data.\n")
        } 
        next
      } else {
        locations.train <- data.table::rbindlist(
          list(
            data.table(x=Q$data$x, y=Q$data$y, obs=T), 
            data.table(x=Q$dummy$x, y=Q$dummy$y, obs=F)
          )
        ) 
        if (covariates.keep > 0) {
          covariates.rasters <- names(covariates) %>%
            set_names() %>%
            purrr::map(function(x) {
              r <- terra::rast(covariates[[x]])
              set.crs(r, st_crs(4326, parameters=T)$Wkt)
              names(r) <- x
              r
            })
          purrr::walk2(names(covariates), covariates.rasters, function(n, r) {
            locations.train[, (n) := 
                             terra::extract(r, cbind(locations.train$x, 
                                                     locations.train$y))]
          })
        }
        glm.pred <- spatstat.model::predict.ppm(
          fit.glm, 
          locations=locations.train[, .(x,y)],
          type="trend", se=T)
        # Intensity
        inten <- predict(fit.glm)
        pixarea <- with(inten, xstep * ystep)
        glm.train <- cbind(
          locations.train,
          data.table(
            est=glm.pred$estimate,
            se=glm.pred$se,
            p.obs = glm.pred$estimate*pixarea)
        )
        optimal.threshold <- optimize.f1(glm.train)
        cm <- get.acc(glm.train, optimal.threshold)
        acc <- tibble(
          common.name=spec,
          state=st,
          covariate.count=covariates.keep,
          optimal.threshold=optimal.threshold 
        ) %>%
          cbind(as.list(c(cm$overall, cm$byClass)) %>% 
                  as_tibble()) %>%
        select(common.name:Accuracy, Sensitivity, Specificity, F1)
        cat("Train Results:\n Species:", spec, "\n",
            "State:", st, "\n",
            "Covariates:", covariates.keep, "\n",
            "Optimal Threshold:", optimal.threshold, "\n",
            "Accuracy:", acc$Accuracy, "\n",
            "F1:", acc$F1, "\n",
            "Sensitivity (TP Rate):", acc$Sensitivity, "\n",
            "Specificity (TN Rate):", acc$Specificity, "\n")
        if ((acc$Sensitivity == 1 & acc$Specificity == 0) | 
             (acc$Sensitivity == 0 & acc$Specificity == 1)) {
          cat("\tThe sensitivity/specificity is a 0/1 pair",
              "for covariates.keep ==", covariates.keep, "\n")
          spec.sens.check <- F
          file.remove(glm.path)
          if (covariates.keep > 15) {
            covariates.keep <- covariates.keep - 5
          } else {
            covariates.keep <- covariates.keep - 1
          }
          if (covariates.keep < 3 & !alternate.formula) {
            cat("\tNo trend identified. Removing interactions...\n")
            alternate.formula <- T
            covariates.keep <- 50
          }  else if (covariates.keep == 0 & alternate.formula) {
            cat("\tNo trend identified. Removing covariates...\n")
          } else if (covariates.keep < 0 & alternate.formula) {
            cat("\tNo other meaningful options to check, 
                   assuming the model has no trend;\n")
            spec.sens.check <- T
          } 
          next
        } else {
          spec.sens.check <- T
        }
      }
    }
    
    glm.results <- get.object(
      obj={
        locations.test <- data.table::rbindlist(
          list(
            data.table(x=Q.test$data$x, y=Q.test$data$y, obs=T), 
            data.table(x=Q.test$dummy$x, y=Q.test$dummy$y, obs=F)
          )
        ) 
        if (covariates.keep > 0) {
          covariates.rasters <- names(covariates) %>%
            set_names() %>%
            purrr::map(function(x) {
              r <- terra::rast(covariates[[x]])
              set.crs(r, st_crs(4326, parameters=T)$Wkt)
              names(r) <- x
              r
            })
          purrr::walk2(names(covariates), covariates.rasters, function(n, r) {
            locations.test[, (n) := 
                             terra::extract(r, cbind(locations.test$x, 
                                                     locations.test$y))]
          })
        }
        glm.pred <- spatstat.model::predict.ppm(
          fit.glm, 
          locations=locations.test[, .(x,y)],
          type="trend", se=T)
        glm.ci <- spatstat.model::predict.ppm(
          fit.glm, 
          locations=locations.test[, .(x,y)],
          type="trend", interval="c")
        glm.test <- cbind(
          locations.test,
          data.table(
            est=glm.pred$estimate,
            se=glm.pred$se,
            lo=glm.ci[1:length(glm.pred$estimate)],
            hi=glm.ci[(length(glm.pred$estimate) + 1):length(glm.ci)])
        )
        
        glm.test[, `:=` (
          p.obs = est * pixarea, # probability > 0
          p.obs.lo = est * lo,
          p.obs.hi = est * hi
        )]
        cm <- get.acc(glm.test, optimal.threshold)
        test.acc <- tibble(
          common.name=spec,
          state=st,
          covariate.count=covariates.keep,
          optimal.threshold=optimal.threshold 
        ) %>%
          cbind(as.list(c(cm$overall, cm$byClass)) %>% 
                  as_tibble()) %>%
          select(common.name:Accuracy, Sensitivity, Specificity, F1)
        all.predictions <- response.ppm(fit.glm)$window %>% as.matrix()
        glm.count.pred <- tryCatch(spatstat.model::predict.ppm(fit.glm, 
                                                               type="count", se=T),
                           error=function(e) NULL)
        glm.pi <- tryCatch(spatstat.model::predict.ppm(fit.glm, type="count", 
                                                       se=T, interval="p"),
                           error=function(e) NULL)
        list(
          test=glm.test,
          train=glm.train,
          all.preds=all.predictions,
          count=glm.count.pred,
          pi=glm.pi,
          thresh=optimal.threshold,
          train.accuracy=acc,
          test.accuracy=test.acc
        )
      },
      file.name=paste0(spec, "_", st, "_ipp_glm_mpl_2.rds"),
      obj.path="artifacts/test_results/ipp_glm_mpl_2"
    )
    cat("\tFinished IPP model for", spec, "in", st, "\n")
  }
  gc()
})

```


```{r examine-mel-results}

ipp.models <- purrr::map_df(1:nrow(spec.state), function(i) {
  spec <- spec.state[i,]$species
  st <- spec.state[i,]$state
  results.path <- file.path("artifacts/test_results/ipp_glm_mpl_2",
                        paste0(spec, "_", st, "_ipp_glm_mpl_2.rds"))
  readRDS(results.path)$test.accuracy
})
ipp.models %>% 
  filter(Sensitivity == 0 | Specificity == 0) %>%
  select(common.name, state, optimal.threshold,
         Sensitivity, Specificity) %>%
  print(n=32)
ipp.models

```

```{r model-acc}
summary(ipp.models)
```


```{r box-plots-1, fig.width=4, fig.height=5}

# Assuming your dataframe is named `ipp.models`
# Reshape the data to long format
long.data <- ipp.models %>%
  select(common.name, Accuracy:F1) %>%
  gather(key = "Metric", value = "Value", -common.name)

# Overall box-plot representation
ggplot(long.data, aes(x = Metric, y = Value)) +
  geom_boxplot(fill = "white", outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, color = "red") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "lightgray"),
        plot.background = element_rect(fill = "white"),
        legend.position = "none") +
  coord_flip()

```

```{r box-plot-2, fig.width=8, fig.height=20}

# Box plot for each species
ggplot(long.data, aes(x = Metric, y = Value)) +
  geom_boxplot(fill = "white", outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, color = "red") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "lightgray"),
        plot.background = element_rect(fill = "white"),
        legend.position = "none") +
  coord_flip() +
  facet_wrap(~ common.name, scales = "free_y", ncol = 2)

```


<br style="margin:0; height:0; padding:0;">


